---
title: "Data Processing and Visualization for the Global AMR Gender Study"
format: html
editor: visual
---

# Introduction

This document contains the analysis and visualization steps for reproducing Figure 5 (Model 1), along with visualizations for the other model variants.

All required preprocessing and model fitting steps are executed via a wrapper script called `main.R`. This script sources the following R scripts in the correct order:

-   `data.R` – data preparation

-   `funcs.R` – utility functions

-   `model.R` – statistical model fitting (computationally intensive)

-   `ComparisonModelLognormal.R` and `ComparisonModelGaussian.R` – comparison of log-normal vs. log-transformed normal models (computationally intensive)

-   `StratifiedModelARG.R` and `StratifiedModelShannon.R` – stratified model fitting by subgroups such as income group and gender (computationally intensive)

-   `HierarchicalModelARG.R` and `HierarchicalModelShannon.R` – hierarchical modeling with nested random effects (computationally intensive)

-   `FrequentistCompareARG.R` and `FrequentistCompareShannon.R` – simulation-based comparison of Bayesian and frequentist models (computationally intensive)

-   `EnrichmentModels.R` – runs enrichment models using Bayesian regression to estimate high ARG prevalence across groups (computationally very intensive)

- `Stancompare.R` – comparison between rtsan and cmdstanr (computationally very intensive)

Before rendering this report, ensure that all these scripts are available in your working directory. To render this report and run all required scripts, make sure to execute script `main.R`.

The dataset (`tse_AMRdemo.rds`) used in this analysis, as well as all the above scripts, can be found in the [GitHub](https://github.com/microbiome/data/blob/main/Salehi2025). This separation ensures that model fitting, which is computationally intensive, does not need to be rerun every time the report is generated.

```{r setup, include=FALSE}
# Load all the necessary libraries
library(scater)
library(ggpubr)
library(rstatix)
library(viridis)
library(ggsignif)
library(Matrix)
library(RColorBrewer)
library(caret)
library(randomForest)
library(xgboost)
library(brms)
library(glmnet)
library(tidyverse)
library(vegan)
library(TreeSummarizedExperiment)
library(dplyr)
library(loo)
library(cmdstanr)
library(broom)
library(tibble)
library(tictoc)
```

# Data processing

All preprocessing steps including filtering samples, extracting metadata, calculating resistance class abundances, and computing diversity metrics are implemented in a separate R script (`data.R`). Please ensure that this script is located in your working directory or adjust the path accordingly.

# Utility functions

All helper functions such as formatting numbers and assigning significance indicators are implemented in a separate R script (`funcs.R`). Please ensure that this script is available in your working directory.

# Modeling

This section summarizes the results from four Bayesian regression model variants. Each model was fitted separately for two outcomes ARG load and Shannon diversity and stratified by income group: HIC (High-Income Countries) and LMIC (Low- and Middle-Income Countries).

### Model Variants:

1.  **Model 1 (Reference):** Fixed effects only (no adjustment for bioproject or sequencing depth):

    $$
    y_i \sim \mathcal{D}(\mu_i), \quad \mu_i = \beta_0 + \mathbf{X}_i \boldsymbol{\beta}
    $$

2.  **Model 2:** Adds a random intercept for `bioproject` to account for study-level variability:

    $$
    y_i \sim \mathcal{D}(\mu_i), \quad \mu_i = \beta_0 + \mathbf{X}_i \boldsymbol{\beta} + b_{\text{project}[i]}
    $$

3.  **Model 3:** Includes sequencing depth as a covariate (log-transformed total read count):

    $$
    y_i \sim \mathcal{D}(\mu_i), \quad \mu_i = \beta_0 + \mathbf{X}_i \boldsymbol{\beta} + \beta_{\text{depth}} \cdot \log(\text{readcount}_i)
    $$

4.  **Model 4:** Combines both random intercept and sequencing depth:

    $$
    y_i \sim \mathcal{D}(\mu_i), \quad \mu_i = \beta_0 + \mathbf{X}_i \boldsymbol{\beta} + \beta_{\text{depth}} \cdot \log(\text{readcount}_i) + b_{\text{project}[i]}
    $$

Where:

-   $y_i$ is the outcome (either ARG load or Shannon diversity),
-   $\mathcal{D}$ is the distribution (log-normal for ARG load, Gaussian for Shannon),
-   $\mathbf{X}_i$ are the dummy-encoded covariates (e.g., gender, age group, region, antibiotic use, etc.),
-   $\boldsymbol{\beta}$: vector of fixed effect coefficients,
-   $b_{\text{project}[i]}$ is the random intercept for bioproject.

Model fitting is computationally intensive and has been performed in a separate R script (`model.R`). Please ensure that this script is available in your working directory.

The resulting model objects are saved as `.rds` files for each model (e.g., `fit1_ARG_load.rds`, `fit1_shannon_diversity.rds`, etc.). These files are then loaded into this document to allow downstream summarization and visualization without re-running the computationally intensive model fitting.

The models are then combined into respective lists (`fit_list1`, `fit_list2`, `fit_list3`, and `fit_list4`) for easier manipulation and analysis in subsequent steps.

```{r, load_models, include=FALSE}
# Load models
fit1_ARG_load <- readRDS("fit1_ARG_load.rds")
fit1_shannon_diversity <- readRDS("fit1_shannon_diversity.rds")
fit2_ARG_load <- readRDS("fit2_ARG_load.rds")
fit2_shannon_diversity <- readRDS("fit2_shannon_diversity.rds")
fit3_ARG_load <- readRDS("fit3_ARG_load.rds")
fit3_shannon_diversity <- readRDS("fit3_shannon_diversity.rds")
fit4_ARG_load <- readRDS("fit4_ARG_load.rds")
fit4_shannon_diversity <- readRDS("fit4_shannon_diversity.rds")

# Combine models into respective lists
fit_list1 <- list(
  fit1_ARG_load = fit1_ARG_load,
  fit1_shannon_diversity = fit1_shannon_diversity
)

fit_list2 <- list(
  fit2_ARG_load = fit2_ARG_load,
  fit2_shannon_diversity = fit2_shannon_diversity
)

fit_list3 <- list(
  fit3_ARG_load = fit3_ARG_load,
  fit3_shannon_diversity = fit3_shannon_diversity
)

fit_list4 <- list(
  fit4_ARG_load = fit4_ARG_load,
  fit4_shannon_diversity = fit4_shannon_diversity
)
```

Then, we define the response variables (ARG_load, shannon_diversity) and income groups (LMIC, HIC) used for stratified statistical modeling.

```{r}
incomes <- c("LMIC", "HIC")
set.seed(123123)
```

For each of the 4 model variants, we extract posterior estimates (mean and 95% credible interval) from the brms model fits. This is done separately for:

-   two response variables: ARG load and Shannon diversity

-   two income groups: HIC and LMIC

For each model, we extract coefficient estimates, 95% credible intervals, and compute whether the effect is statistically significant.

```{r}
# Results from Model 1: Fixed effects only (Reference)
full_summary1 <- lapply(names(fit_list1), function(r) {  
  lapply(incomes, function(ic) {
    
    # Extract posterior summaries for fixed effects (only coefficients for fixed effects)
    my_summary <- posterior_summary(fit_list1[[r]][[ic]]) %>%
      .[grep("^b_", rownames(.)), ] %>%
      as.data.frame() %>%
      rownames_to_column(var = "Feature") %>%
      mutate(
        Feature = gsub("b_", "", Feature),
        Response = gsub("^fit[0-9]+_", "", r),
        income_group = ic,
        Significant = ifelse(Q2.5 > 0 | Q97.5 < 0, TRUE, FALSE)  
      ) %>%
      dplyr::select(-Est.Error)
    
    return(my_summary)
  }) %>% do.call(rbind, .)
}) %>% do.call(rbind, .) %>% mutate(model = 1)


# Results from Model 2: + Random intercept for bioproject
full_summary2 <- lapply(names(fit_list2), function(r) {
  lapply(incomes, function(ic) {
    
    # Posterior summaries including adjustment for study-level random effects
    my_summary <- posterior_summary(fit_list2[[r]][[ic]]) %>%
      .[grep("^b_", rownames(.)), ] %>%
      as.data.frame() %>%
      rownames_to_column(var = "Feature") %>%
      mutate(
        Feature = gsub("b_", "", Feature),
        Response = gsub("^fit[0-9]+_", "", r),
        income_group = ic,
        Significant = ifelse(Q2.5 > 0 | Q97.5 < 0, TRUE, FALSE)
      ) %>%
      dplyr::select(-Est.Error)
    
    return(my_summary)
  }) %>% do.call(rbind, .)
}) %>% do.call(rbind, .) %>% mutate(model = 2)


# Results from Model 3: + log(readcount) as covariate
full_summary3 <- lapply(names(fit_list3), function(r) {
  lapply(incomes, function(ic) {
    
    # Posterior summaries including adjustment for sequencing depth
    my_summary <- posterior_summary(fit_list3[[r]][[ic]]) %>%
      .[grep("^b_", rownames(.)), ] %>%
      as.data.frame() %>%
      rownames_to_column(var = "Feature") %>%
      mutate(
        Feature = gsub("b_", "", Feature),
        Response = gsub("^fit[0-9]+_", "", r),
        income_group = ic,
        Significant = ifelse(Q2.5 > 0 | Q97.5 < 0, TRUE, FALSE)
      ) %>%
      dplyr::select(-Est.Error)
    
    return(my_summary)
  }) %>% do.call(rbind, .)
}) %>% do.call(rbind, .) %>% mutate(model = 3)


# Results from Model 4: + log(readcount) + (1 | bioproject)
full_summary4 <- lapply(names(fit_list4), function(r) {
  lapply(incomes, function(ic) {
    
    # Posterior summaries including both sequencing depth and study-level random effect
    my_summary <- posterior_summary(fit_list4[[r]][[ic]]) %>%
      .[grep("^b_", rownames(.)), ] %>%
      as.data.frame() %>%
      rownames_to_column(var = "Feature") %>%
      mutate(
        Feature = gsub("b_", "", Feature),
        Response = gsub("^fit[0-9]+_", "", r),
        income_group = ic,
        Significant = ifelse(Q2.5 > 0 | Q97.5 < 0, TRUE, FALSE)
      ) %>%
      dplyr::select(-Est.Error)
    
    return(my_summary)
  }) %>% do.call(rbind, .)
}) %>% do.call(rbind, .) %>% mutate(model = 4)

```

# Combining results

This section combines the outputs from all four Bayesian regression models into a unified summary table. Each model's results include the estimated effect sizes and confidence intervals for all predictors. The summary table is cleaned, standardized, and formatted for further analysis and plotting.

```{r, combine}
# Combine all model summaries into a single data frame
full_summary <- list(
  full_summary1, 
  full_summary2, 
  full_summary3, 
  full_summary4
) %>% 
  do.call(rbind, .)

# Recode and relabel model names for clarity
full_summary <- full_summary %>%
  mutate(
    model = recode(as.character(model),
                   "1" = "original", 
                   "2" = "+(1|bioproject)", 
                   "3" = "+log(readcount)", 
                   "4" = "+(1|bioproject)+log(readcount)"),
    model = factor(model, levels = c(
      "original", 
      "+(1|bioproject)", 
      "+log(readcount)", 
      "+(1|bioproject)+log(readcount)"))
  )

# Clean and harmonize variable names for better readability
full_summary <- full_summary %>%
  mutate(
    Feature = gsub("^continent", "", Feature),
    Feature = gsub("^age_category", "", Feature),
    Feature = gsub("^genderWomen", "Woman", Feature),
    Feature = gsub("^Usage_high1", "High Antibiotic Use", Feature),
    Feature = gsub("([a-z])([A-Z])", "\\1 \\2", Feature),
    Feature = gsub("_", " ", Feature),
    Feature = trimws(Feature),
    Feature = gsub("logreadcount", "log(read count)", Feature),
    Response = gsub("ARG_load", "ARG load", Response),
    Response = gsub("shannon_diversity", "Shannon", Response)
  )

# Select and rename relevant columns
full_summary <- full_summary %>%
  select(
    Predictor = Feature,
    Estimate, Q2.5, Q97.5,
    Response,
    `Income Group` = income_group, 
    model
  )

# Calculate exponentiated estimates and confidence intervals
full_summary <- full_summary %>%
  mutate(
    `exp(Estimate)` = exp(Estimate),
    `exp(Q2.5)` = exp(Q2.5),
    `exp(Q97.5)` = exp(Q97.5)
  )

# Define display order for predictors in plots
full_summary$Predictor <- factor(full_summary$Predictor,
  levels = c(
    "log(read count)", 
    "South America", "Oceania", "North America", "Europe", "Asia",
    "Oldest Adult", "Older Adult", "Young Adult", "Teenager",
    "Children", "Toddler", "Infant",
    "High Antibiotic Use", "Woman", "Intercept"
  )
)

# Rename model column and update label for reference model
full_summary <- full_summary %>%
  rename(Model = model) %>%
  mutate(Model = recode(Model, "original" = "Reference"))
```

# Model Results

In this section, detailed results are presented for each of the four Bayesian regression models introduced earlier. For every model, we provide both a summary table of posterior estimates and a corresponding plot visualizing the effect sizes and 95% credible intervals for each predictor. Results are shown separately for the two outcome variables (ARG load and Shannon diversity) and stratified by income group (HIC and LMIC). A combined comparison plot is included at the end to facilitate visual comparison across models.

## Model 1: Fixed Effects Only

```{r}
# Filter Model 1 results from full_summary
model1_summary <- full_summary %>%
  filter(Model == "Reference") %>%
  mutate(
    lower = (exp(Q2.5) - 1) * 100,
    upper = (exp(Q97.5) - 1) * 100,
    mean = (exp(Estimate) - 1) * 100,
    Significant = Q2.5 > 0 | Q97.5 < 0
  ) %>%
  filter(Predictor != "Intercept")

# Summary table
model1_summary %>%
  dplyr::select(Predictor, Response, `Income Group`, Estimate, Q2.5, Q97.5, Significant) %>%
  knitr::kable(digits = 3, caption = "Posterior summaries for Model 1 (Fixed Effects Only)")

# Plot model 1
p1 <- full_summary %>% 
  filter(Model == levels(full_summary$Model)[1]) %>%
  mutate(lower = (exp(Q2.5) - 1)*100,
         upper = (exp(Q97.5) - 1)*100, 
         mean = (exp(Estimate) - 1)*100) %>%
  filter(Predictor != "Intercept") %>% 
  ggplot() + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(x = Predictor, ymin = lower, ymax = upper, color = Response),
                position = position_dodge(width = 0.5), width = 0.4, linewidth = 1) +
  geom_point(aes(x = Predictor, y = (exp(Estimate) - 1)*100,
                 color = Response),
             position = position_dodge(width = 0.5),
             size = 2) +
  facet_wrap(~`Income Group`) +
  coord_flip() +
  labs(title = "Model 1: Fixed Effects Only", y = "Effect Size (%)", x = "") +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e")) +
  facet_wrap(~`Income Group`, ncol = 2) + 
  theme_bw(11) + 
  theme(strip.background =element_rect(fill="white"))

p1
```

## Model 2: + (1 \| bioproject)

```{r}
# Filter Model 2 results from full_summary
model2_summary <- full_summary %>%
  filter(Model == "+(1|bioproject)") %>%
  mutate(
    lower = (exp(Q2.5) - 1) * 100,
    upper = (exp(Q97.5) - 1) * 100,
    mean = (exp(Estimate) - 1) * 100,
    Significant = Q2.5 > 0 | Q97.5 < 0
  ) %>%
  filter(Predictor != "Intercept")

# Summary table
model2_summary %>%
  dplyr::select(Predictor, Response, `Income Group`, Estimate, Q2.5, Q97.5, Significant) %>%
  knitr::kable(digits = 3, caption = "Posterior summaries for Model 2 (Random Intercept for Bioproject)")

# Plot model 2
p2 <- full_summary %>% 
  filter(Model == levels(full_summary$Model)[2]) %>%
  mutate(lower = (exp(Q2.5) - 1)*100,
         upper = (exp(Q97.5) - 1)*100, 
         mean = (exp(Estimate) - 1)*100) %>%
  filter(Predictor != "Intercept") %>% 
  ggplot() + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(x = Predictor, ymin = lower, ymax = upper, color = Response),
                position = position_dodge(width = 0.5), width = 0.4, linewidth = 1) +
  geom_point(aes(x = Predictor, y = (exp(Estimate) - 1)*100,
                 color = Response),
             position = position_dodge(width = 0.5),
             size = 2) +
  facet_wrap(~`Income Group`) +
  coord_flip() +
  labs(title = "Model 2: + (1 | bioproject)", y = "Effect Size (%)", x = "") +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e")) +
  facet_wrap(~`Income Group`, ncol = 2) + 
  theme_bw(11) + 
  theme(strip.background =element_rect(fill="white"))

p2
```

## Model 3: + log(readcount)

```{r}
# Filter Model 3 results from full_summary
model3_summary <- full_summary %>%
  filter(Model == "+log(readcount)") %>%
  mutate(
    lower = (exp(Q2.5) - 1) * 100,
    upper = (exp(Q97.5) - 1) * 100,
    mean = (exp(Estimate) - 1) * 100,
    Significant = Q2.5 > 0 | Q97.5 < 0
  ) %>%
  filter(Predictor != "Intercept")

# Summary table
model3_summary %>%
  dplyr::select(Predictor, Response, `Income Group`, Estimate, Q2.5, Q97.5, Significant) %>%
  knitr::kable(digits = 3, caption = "Posterior summaries for Model 3 (+ log(readcount))")

# Plot model 3
p3 <- full_summary %>% 
  filter(Model == levels(full_summary$Model)[3]) %>%
  mutate(lower = (exp(Q2.5) - 1)*100,
         upper = (exp(Q97.5) - 1)*100, 
         mean = (exp(Estimate) - 1)*100) %>%
  filter(Predictor != "Intercept") %>% 
  ggplot() + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(x = Predictor, ymin = lower, ymax = upper, color = Response),
                position = position_dodge(width = 0.5), width = 0.4, linewidth = 1) +
  geom_point(aes(x = Predictor, y = (exp(Estimate) - 1)*100,
                 color = Response),
             position = position_dodge(width = 0.5),
             size = 2) +
  facet_wrap(~`Income Group`) +
  coord_flip() +
  labs(title = "Model 3: + log(readcount)", y = "Effect Size (%)", x = "") +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e")) +
  facet_wrap(~`Income Group`, ncol = 2) + 
  theme_bw(11) + 
  theme(strip.background =element_rect(fill="white"))

p3
```

## Model 4: + (1 \| bioproject) + log(readcount)

```{r}
# Filter Model 4 results from full_summary
model4_summary <- full_summary %>%
  filter(Model == "+(1|bioproject)+log(readcount)") %>%
  mutate(
    lower = (exp(Q2.5) - 1) * 100,
    upper = (exp(Q97.5) - 1) * 100,
    mean = (exp(Estimate) - 1) * 100,
    Significant = Q2.5 > 0 | Q97.5 < 0
  ) %>%
  filter(Predictor != "Intercept")

# Summary table
model4_summary %>%
  dplyr::select(Predictor, Response, `Income Group`, Estimate, Q2.5, Q97.5, Significant) %>%
  knitr::kable(digits = 3, caption = "Posterior summaries for Model 4 (+ log(readcount) + (1 | bioproject))")

# Plot model 4
p4 <- full_summary %>% 
  filter(Model == levels(full_summary$Model)[4]) %>%
  mutate(lower = (exp(Q2.5) - 1)*100,
         upper = (exp(Q97.5) - 1)*100, 
         mean = (exp(Estimate) - 1)*100) %>%
  filter(Predictor != "Intercept") %>% 
  ggplot() + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(x = Predictor, ymin = lower, ymax = upper, color = Response),
                position = position_dodge(width = 0.5), width = 0.4, linewidth = 1) +
  geom_point(aes(x = Predictor, y = (exp(Estimate) - 1)*100,
                 color = Response),
             position = position_dodge(width = 0.5),
             size = 2) +
  facet_wrap(~`Income Group`) +
  coord_flip() +
  labs( title = "Model 4: + (1 | bioproject) + log(readcount)", y = "Effect Size (%)", x = "") +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e")) +
  facet_wrap(~`Income Group`, ncol = 2) + 
  theme_bw(11) + 
  theme(strip.background =element_rect(fill="white"))

p4

```

## Comparison Across Models

```{r, combineplot, fig.width=14, fig.height=10}

# Create comparison plot across all four models
comparison_p <- full_summary %>% 
  filter(Predictor != "Intercept") %>%
  mutate(
    lower = (exp(Q2.5) - 1) * 100,
    upper = (exp(Q97.5) - 1) * 100,
    mean = (exp(Estimate) - 1) * 100
  ) %>%
  ggplot() + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(
    aes(x = Predictor, ymin = lower, ymax = upper, color = Response),
    position = position_dodge(width = 0.6),
    width = 1,
    linewidth = 1
  ) +
  geom_point(
    aes(x = Predictor, y = mean, color = Response),
    position = position_dodge(width = 0.5),
    size = 1
  ) +
  coord_flip() +
  labs(y = "Effect Size (%)", x = "", title = "Comparison of Effect Estimates Across Models"
  ) +
  facet_wrap(Model ~ `Income Group`, ncol = 2) +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e")) +
  theme_bw(10) +
  theme(
    strip.background = element_rect(fill = "white"),
    axis.text.y = element_text(size = 8)
  )

print(comparison_p)

```

# Comparison of Log-normal and Log-transformed Normal Models

In this analysis, we aim to compare two different statistical modeling approaches for handling skewed data: the log-normal distribution model and the normal distribution model applied to log-transformed data.

1.  **Log-Normal Model**: This model assumes that the data follows a log-normal distribution, which is appropriate when the data is positively skewed. A log-normal distribution is often used when modeling non-negative data with a right-skewed distribution, as it models the logarithm of the data as normally distributed.

2.  **Normal Model on Log-transformed Data**: In this approach, we first apply a log-transformation to the original data (log(x)) and then model the transformed data using a normal distribution. This approach is commonly used when data exhibits skewness, and it can help normalize the data for more effective modeling using a Gaussian distribution.

## Model fitting

Model fitting is computationally intensive and has been performed in a separate R scripts (`ComparisonModelLognormal.R` and `ComparisonModelGaussian.R`). Please ensure that this script is available in your working directory.

The resulting model objects are saved as .rds files for each model (e.g., `fit_ARG_lognormal.rds`, `fit_ARG_gaussian.rds`). These files are then loaded into this document to allow downstream summarization and visualization without re-running the computationally intensive model fitting.

```{r, load2, include=FALSE}
# Load the models
fit_ARG_lognormal <- readRDS("fit_ARG_lognormal.rds")
fit_ARG_gaussian <- readRDS("fit_ARG_gaussian.rds")

# Combine models into respective lists
fit_list_lognormal <- list(
  fit_ARG_load = fit_ARG_lognormal
)

fit_list_gaussian <- list(
  fit_ARG_load = fit_ARG_gaussian
)
```

## Model Comparison

While both models analyze the same outcome conceptually, they are fitted to different versions of the response variable: the log-normal model to the original scale and the Gaussian model to the log-transformed scale (`log(y)`). Because of this, they operate on different likelihoods and data scales.

As a result, information criteria such as WAIC or PSIS-LOO cannot be used to compare them meaningfully. These methods assume models are fitted to the same observed data, which is not the case here. Any visual comparisons presented are therefore illustrative only, and not intended for formal model selection or performance ranking.

To prepare for visualization, we extract and combine the posterior summaries from both models across income groups.

```{r, summary2}
# Results from Model 1: Gaussian with log transform
full_summary_gaussian <- lapply(names(fit_list_gaussian), function(r) {  
  lapply(incomes, function(ic) {
    
    # Extract posterior summaries for fixed effects (only coefficients for fixed effects)
    my_summary <- posterior_summary(fit_list_gaussian[[r]][[ic]]) %>%
      .[grep("^b_", rownames(.)), ] %>%                            
      as.data.frame() %>%
      rownames_to_column(var = "Feature") %>%
      mutate(
        Feature = gsub("b_", "", Feature),                         
        Response = gsub("^fit[0-9]+_", "", r),                     
        income_group = ic,                                         
        Significant = ifelse(Q2.5 > 0 | Q97.5 < 0, TRUE, FALSE)    
      ) %>%
      dplyr::select(-Est.Error)                                    
    
    return(my_summary)
  }) %>% do.call(rbind, .)
}) %>% do.call(rbind, .) %>% mutate(model = "Gaussian")            


# Results from Model 2: Log-normal + random intercept for bioproject
full_summary_lognormal <- lapply(names(fit_list_lognormal), function(r) {
  lapply(incomes, function(ic) {
    
    # Extract posterior summaries including adjustment for study-level random effects
    my_summary <- posterior_summary(fit_list_lognormal[[r]][[ic]]) %>%
      .[grep("^b_", rownames(.)), ] %>%                            
      as.data.frame() %>%
      rownames_to_column(var = "Feature") %>%
      mutate(
        Feature = gsub("b_", "", Feature),                         
        Response = gsub("^fit[0-9]+_", "", r),                     
        income_group = ic,                                         
        Significant = ifelse(Q2.5 > 0 | Q97.5 < 0, TRUE, FALSE)    
      ) %>%
      dplyr::select(-Est.Error)                                    
    
    return(my_summary)
  }) %>% do.call(rbind, .)
}) %>% do.call(rbind, .) %>% mutate(model = "Log-normal") 

# Combine results from both models
full_summary <- list(
  full_summary_lognormal,
  full_summary_gaussian
) %>% do.call(rbind, .)

# Format model labels as a factor with custom order
full_summary <- full_summary %>%
  mutate(
    model = factor(model, levels = c("Log-normal", "Gaussian"))
  )

# Clean and harmonize feature and response names
full_summary <- full_summary %>%
  mutate(
    Feature = gsub("^continent", "", Feature),
    Feature = gsub("^age_category", "", Feature),
    Feature = gsub("^genderWomen", "Woman", Feature),
    Feature = gsub("^Usage_high1", "High Antibiotic Use", Feature),
    Feature = gsub("([a-z])([A-Z])", "\\1 \\2", Feature),
    Feature = gsub("_", " ", Feature),
    Feature = trimws(Feature),
    Feature = gsub("logreadcount", "log(read count)", Feature),
    Response = gsub("ARG_load", "ARG load", Response),
    Response = gsub("shannon_diversity", "Shannon", Response)
  )

# Rename and select relevant columns for summary table and plotting
full_summary <- full_summary %>%
  select(
    Predictor = Feature,
    Estimate, Q2.5, Q97.5,
    Response,
    `Income Group` = income_group, 
    model
  )

# Compute exponentiated estimates and confidence intervals for interpretation
full_summary <- full_summary %>%
  mutate(
    `exp(Estimate)` = exp(Estimate),
    `exp(Q2.5)` = exp(Q2.5),
    `exp(Q97.5)` = exp(Q97.5)
  )

# Set custom display order for predictors
full_summary$Predictor <- factor(full_summary$Predictor,
  levels = c(
    "log(read count)", 
    "South America", "Oceania", "North America", "Europe", "Asia",
    "Oldest Adult", "Older Adult", "Young Adult", "Teenager",
    "Children", "Toddler", "Infant",
    "High Antibiotic Use", "Woman", "Intercept"
  )
)

```

To explore potential differences between the models, we next visualize the estimated effects across predictors and income groups.

```{r, combineplot2, fig.width=14, fig.height=10}
comparison_log <- full_summary %>%
  filter(Predictor != "Intercept") %>%
  mutate(
    lower = (exp(Q2.5) - 1) * 100,
    upper = (exp(Q97.5) - 1) * 100,
    mean = (exp(Estimate) - 1) * 100
  ) %>%
  ggplot() + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  
  geom_errorbar(aes(x = Predictor, ymin = lower, ymax = upper),
                position = position_dodge(width = 0.6), width = 0.5, linewidth = 1,
                color = "#1f77b4"
  ) +
  
  geom_point(aes(x = Predictor, y = mean), position = position_dodge(width = 0.6),
             size = 1.5, color = "#1f77b4"
  ) +
  
  coord_flip() +
  labs(y = "Effect Size (%)",  x = "",  
       title = "Comparison of Effect Estimates Across Models"
  ) +
  facet_wrap(model ~ `Income Group`, ncol = 2) +
  theme_bw(14) +
  theme(strip.background = element_rect(fill = "white"), 
        axis.text.y = element_text(size = 12)
  )

print(comparison_log)
```

Based on the visualization, there are no substantial differences in the estimated effects between the two models across predictors and income groups. In terms of computational performance, model fitting times were also very similar, providing no practical advantage to either approach.

Taken together, these observations do not favor one model over the other based on empirical results. However, from a modeling perspective, the log-normal model remains the more principled choice, as it directly models the original skewed data without requiring transformation.

# Age Effects Stratified by Income Group and Gender

To explore how age affects microbial outcomes across different populations, we fitted separate models within four demographic subgroups defined by income group (HIC vs. LMIC) and gender (Women vs. Men). Each model includes only age category as a predictor, allowing us to examine age-related patterns while holding other factors constant.

This stratified approach provides a simple way to evaluate whether age-related effects differ between, for example, women in high-income countries and men in low- and middle-income countries. The modeling was performed separately for both outcome variables: ARG load and Shannon diversity.

We fit separate models for each combination of income group and gender. Each model includes only age category as a fixed effect, and no additional covariates.

The general model is:

$$
y_i \sim \begin{cases}\text{LogNormal}(\mu_i, \sigma^2) & \text{for ARG load} \\\\\mathcal{N}(\mu_i, \sigma^2) & \text{for Shannon diversity}\end{cases}
$$

$$
\mu_i = \beta_0 + \sum_l \beta_l \cdot \text{AgeCategory}_{il}
$$

Where:

-   $y_i$ is the observed response (e.g., ARG load) for individual $i$
-   $\beta_0$ is the intercept (fixed effect)
-   $\beta_l$ are the fixed effects for each age category $l$
-   $\sigma^2$ is the residual variance

Each model is fit independently for one of the four income–gender subgroups, making no assumptions of shared structure across them.

Model fitting is computationally intensive and has been performed in a separate R scripts (`StratifiedModelARG.R` and `StratifiedModelShannon.R`). Please ensure that this script is available in your working directory. The models have been saved as `.rds` files and are loaded here to enable downstream analysis.

```{r, include=FALSE}
# Load models
fit_ARG_load_age <- readRDS("fit_ARG_load_age.rds")
fit_shannon_diversity_age <- readRDS("fit_shannon_diversity_age.rds")

# Combine models into respective lists
fit_list_age <- list(
  fit_ARG_load_age = fit_ARG_load_age,
  fit_shannon_diversity_age = fit_shannon_diversity_age
)
```

We extract the posterior summaries (posterior means and 95% credible intervals) of fixed-effect coefficients from each model. These include age group and continent effects for each income-gender combination. The results are organized into a single summary table with relevant grouping information (income, gender, and response variable).

```{r}
# Income and gender values
incomes <- c("LMIC", "HIC")
genders <- c("Women", "Men")

# Loop through responses and extract summaries
full_summary_age <- lapply(names(fit_list_age), function(response_name) {
  response_label <- ifelse(response_name == "fit_ARG_load_age", "ARG load", "Shannon")
  fit_list <- fit_list_age[[response_name]]
  
  lapply(incomes, function(ic) {
    lapply(genders, function(g) {
      
      fit <- fit_list[[paste(ic, g, sep = "_")]]
      if (is.null(fit)) return(NULL)
      
      my_summary <- posterior_summary(fit) %>%
        .[grep("^b_", rownames(.)), ] %>%
        as.data.frame() %>%
        rownames_to_column(var = "Feature") %>%
        mutate(
          Feature = gsub("^b_", "", Feature),
          Response = response_label,
          Income = ic,
          Gender = ifelse(g == "Women", "Woman", "Man"),
          Significant = Q2.5 > 0 | Q97.5 < 0
        ) %>%
        dplyr::select(-Est.Error)
      
      return(my_summary)
    }) %>% bind_rows()
  }) %>% bind_rows()
}) %>% bind_rows()

```

To prepare for visualization, we clean the feature names for clarity, compute exponentiated effect sizes, and define a consistent display order for predictors.

```{r}
# Clean and format predictor names and compute exponentiated values
full_summary_age <- full_summary_age %>%
  mutate(
    Predictor = gsub("^age[ _]?category", "", Feature),
    Predictor = recode(Predictor,
                       "YoungAdult" = "Young Adult",
                       "OlderAdult" = "Older Adult",
                       "OldestAdult" = "Oldest Adult"),
    Predictor = trimws(Predictor),
     # Compute exponentiated estimates and credible intervals
    `exp(Estimate)` = exp(Estimate),
    `exp(Q2.5)` = exp(Q2.5),
    `exp(Q97.5)` = exp(Q97.5)
  ) %>%
  select(Predictor, Estimate, Q2.5, Q97.5,
         `exp(Estimate)`, `exp(Q2.5)`, `exp(Q97.5)`,
         Response, Income, Gender, Significant)


# Define desired display order for predictors in plots
all_predictors <- c(
  "Oldest Adult", "Older Adult", "Young Adult", "Teenager",
  "Children", "Toddler", "Infant"
)
full_summary_age$Predictor <- factor(full_summary_age$Predictor, levels = all_predictors)
```

Finally, we plot the estimated effect sizes for each predictor. The plot is faceted by gender and income group to allow for easy comparison of demographic patterns. Both outcomes are shown side by side in color, with error bars representing 95% credible intervals.

```{r}
# Plot
plot_age_effects <- full_summary_age %>%
  filter(Predictor != "Intercept") %>%
  mutate(
    mean = (exp(Estimate) - 1) * 100,
    lower = (exp(Q2.5) - 1) * 100,
    upper = (exp(Q97.5) - 1) * 100
  ) %>%
  ggplot(aes(x = Predictor, y = mean, color = Response)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.3, linewidth = 1.2,
                position = position_dodge(width = 0.5)) +
  geom_point(size = 2,
             position = position_dodge(width = 0.5)) +
  facet_grid(Gender ~ Income) +
  coord_flip() +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e")) +
  labs(
    title = "Effect Estimates by Age and Continent\nStratified by Income Group and Gender",
    y = "Effect Size (%)", x = "", color = "Response"
  ) +
  theme_bw(12) +
  theme(strip.background = element_rect(fill = "white"))

print(plot_age_effects)
```

# Hierarchical Age Model by Income and Gender

We use a hierarchical model that mirrors the structure of the data: income group → gender → age category. People are first grouped by income level, then by gender within each income group, and finally by age category. This setup lets the age effects differ between income and gender groups, making it easier to spot patterns that vary across these groups.

We use either a normal or log-normal likelihood depending on the outcome:

$$
y_{ijk} \sim \begin{cases}\text{LogNormal}(\mu_{ijk}, \sigma^2) & \text{for ARG load} \\\\\mathcal{N}(\mu_{ijk}, \sigma^2) & \text{for Shannon diversity}\end{cases}
$$

$$
\mu_{ijk} = \beta_0 + \sum_l \beta_l \cdot \text{AgeCategory}_{ijkl} + b_{0i} + b_{0ij} + \sum_l b_{li} \cdot \text{AgeCategory}_{ijkl} + \sum_l b_{lij} \cdot \text{AgeCategory}_{ijkl}
$$

Where:

-   $y_{ijk}$ is the response for individual $k$ in gender group $j$ and income group $i$ (e.g., Shannon diversity or ARG load)

-   $\beta_0$ is the overall intercept (fixed effect)

-   $\beta_l$ are the fixed effects for each age category $l$

-   $b_{0i}$ is the random intercept for income group $i$

-   $b_{0ij}$ is the random intercept for gender group $j$ within income group $i$

-   $b_{li}$ are the random slopes for age category $l$ within income group $i$

-   $b_{lij}$ are the random slopes for age category $l$ within gender group $j$ and income group $i$

-   $\sigma^2$ is the residual variance

To reduce computation time and avoid convergence issues, the hierarchical models were fitted on a subset of the full dataset. Specifically, 20 observations were randomly selected from each combination of income group, gender, and age category, ensuring balanced representation across all demographic strata.

Model fitting in this case is computationally intensive and has been performed in a separate R scripts (`HierarchicalModelARG.R` and `HierarchicalModelShannon.R`). Please ensure that this script is available in your working directory. The models have been saved as `.rds` files and are loaded here to enable downstream analysis.

```{r, include=FALSE}
# Load hierarchical models (age only)
# These models were fitted using a multilevel structure:
# (age_category + (1 + age_category | income_group_HIC / gender))

fit_ARG_hierarchical <- readRDS("fit_ARG_hierarchical.rds")
fit_shannon_hierarchical <- readRDS("fit_shannon_hierarchical.rds")

# Combine models into a named list for processing
fit_list_hierarchical <- list(
  "ARG load" = fit_ARG_hierarchical,
  "Shannon" = fit_shannon_hierarchical
)
```

We extract the posterior summaries (posterior means and 95% credible intervals) of fixed-effect coefficients from each model. These include age group and continent effects for each income-gender combination. The results are organized into a single summary table with relevant grouping information (income, gender, and response variable). Also to prepare for visualization, we clean the feature names for clarity, compute exponentiated effect sizes, and define a consistent display order for predictors.

```{r}
# Extract group-specific fixed + random effects
full_summary_hier <- lapply(names(fit_list_hierarchical), function(name) {
  model <- fit_list_hierarchical[[name]]

  posterior_summary(model) %>%
    as.data.frame() %>%
    rownames_to_column("term") %>%
    filter(str_detect(term, "r_income_group_HIC:gender\\[")) %>%
    mutate(
      Response = name,
      # Extract inside brackets, e.g., "0_Women,age_categoryInfant"
      Group = str_extract(term, "\\[.*?\\]") %>% str_remove_all("\\[|\\]"),
      # Split group into income + gender
      Income_raw = str_split(Group, ",", simplify = TRUE)[,1],
      Var_raw = str_split(Group, ",", simplify = TRUE)[,2],
      Income = ifelse(str_detect(Income_raw, "^1"), "HIC", "LMIC"),
      Gender = ifelse(str_detect(Income_raw, "Women"), "Woman", "Man"),
      Predictor = str_remove(Var_raw, "age_category"),
      Predictor = recode(Predictor,
        "YoungAdult" = "Young Adult",
        "OlderAdult" = "Older Adult",
        "OldestAdult" = "Oldest Adult"
      ),
      Predictor = trimws(Predictor)
    ) %>%
    filter(!is.na(Predictor)) %>%
    mutate(
      `exp(Estimate)` = ifelse(Response == "ARG load", exp(Estimate), Estimate),
      `exp(Q2.5)` = ifelse(Response == "ARG load", exp(Q2.5), Q2.5),
      `exp(Q97.5)` = ifelse(Response == "ARG load", exp(Q97.5), Q97.5)
    ) %>%
    select(Predictor, Estimate, Q2.5, Q97.5,
           `exp(Estimate)`, `exp(Q2.5)`, `exp(Q97.5)`,
           Response, Income, Gender)
}) %>% bind_rows()


# Order predictors
full_summary_hier$Predictor <- factor(full_summary_hier$Predictor, levels = all_predictors)
```

Next, we visualize the estimated effects from the hierarchical models.

```{r}
# Plot
plot_hierarchical_age <- full_summary_hier%>%
  filter(!is.na(Predictor), Predictor != "Intercept") %>%
  mutate(
    mean = (`exp(Estimate)` - 1) * 100,
    lower = (`exp(Q2.5)` - 1) * 100,
    upper = (`exp(Q97.5)` - 1) * 100
  ) %>%
  ggplot(aes(x = Predictor, y = mean, color = Response)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.3, linewidth = 1.2,
    position = position_dodge(width = 0.5)
  ) +
  geom_point(
    size = 2,
    position = position_dodge(width = 0.5)
  ) +
  facet_grid(Gender ~ Income) +
  coord_flip() +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e")) +
  labs(
    title = "Effect Estimates by Age Category\nStratified by Income Group and Gender (Hierarchical Model)",
    y = "Effect Size (%)", x = "", color = "Response"
  ) +
  theme_bw(base_size = 12) +
  theme(strip.background = element_rect(fill = "white"))

print(plot_hierarchical_age)
```

This figure shows age-related effect estimates from a hierarchical model stratified by income group and gender. Because fitting the full model on the entire dataset was too computationally intensive, a random subset was used instead with a maximum of 20 observations per group. As a result, the model is not very stable, and some of the estimates appear unrealistically large with wide uncertainty intervals. The subset is not representative enough to support reliable inference. These results highlight the limitations of hierarchical modeling under small sample sizes, and emphasize the need for either simplified models or greater computational resources when using the full dataset.

# Comparing Model Stability Under Small Samples: Probabilistic vs. Frequentist Approaches

This section explores how probabilistic (Bayesian) and frequentist models behave under limited data conditions. The goal is to evaluate model robustness and stability when the available sample size is small.

To achieve this, we repeatedly draw random subsamples of varying sizes from a larger dataset. For each subsample, we fit two models:

-   **A Bayesian model**, implemented with the brms package

-   **A frequentist linear model** using standard linear regression (lm())

We extract the effect estimates and the width of the 95% uncertainty intervals (credible intervals for Bayesian, confidence intervals for frequentist) for age category predictors. By comparing these across repetitions and sample sizes, we assess how each model handles uncertainty under constrained data scenarios.

The comparison is performed for two outcome variables:

1.  ARG load (modeled using a lognormal likelihood)

2.  Shannon diversity index (modeled using a Gaussian likelihood)

Model fitting and repetitions are computationally intensive and have been performed in a separate R scripts (`FrequentistCompareARG.R` and `FrequentistCompareShannon.R`). Please ensure that this script is available in your working directory. The model outputs have been saved as `.rds` files and are loaded here for downstream analysis.

```{r, include=FALSE}
# Load simulation results for both models
results_ARG <- readRDS("results_compare_ARG.rds")
results_shannon <- readRDS("results_compare_shannon.rds")
```

## Results: ARG load

The first analysis focuses on ARG load as the response variable, modeled using a lognormal likelihood. The following visualizations and summary statistics are based on simulation results obtained from repeated subsampling. Five independent subsamples were drawn for each of the selected sample sizes: 30, 80, 150, 250, and 500. For each subsample, both a Bayesian and a frequentist model were fitted to estimate the effects of age categories on ARG load. The focus is on comparing the width and stability of the 95% uncertainty intervals produced by the two modeling approaches across varying sample sizes.

```{r}
# Plot the distribution of 95% interval widths by model and sample size (ARG model)
# Separate panels are shown for each age category
desired_order <- c("Infant", "Toddler", "Children", "Teenager", 
                   "Young Adult", "Older Adult", "Oldest Adult")
results_ARG$Predictor <- factor(results_ARG$Predictor, levels = desired_order)
ggplot(results_ARG, aes(x = factor(sample_size), y = width, fill = model)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.85, linewidth = 0.6) +
  facet_wrap(~Predictor, scales = "free_y", ncol = 3) +
  labs(title = "ARG Load: 95% Interval Width by Sample Size",
       x = "Sample Size", y = "95% Interval Width",
       fill = "Model") +
  scale_fill_manual(values = c("Bayesian" = "green",
                               "Frequentist" = "coral")) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face = "bold"))


# Create a summary table with mean and standard deviation of interval widths
# Grouped by model type and sample size (across all age categories)
summary_table_ARG <- results_ARG %>%
  group_by(model, sample_size) %>%
  summarise(
    mean_width = mean(width),
    sd_width = sd(width),
    .groups = "drop"
  ) %>%
  arrange(sample_size, model)

summary_table_ARG
```

The boxplot illustrates the distribution of 95% interval widths for each age category predictor under both Bayesian and frequentist models. At small sample sizes (e.g., n = 30), Bayesian intervals tend to be narrower and more stable across most age groups. This advantage is particularly evident for categories such as "Oldest Adult", "Toddler", and "Older Adult", where frequentist intervals show greater variability and wider estimates.

As the sample size increases, the two approaches converge toward similar interval widths. By n = 250–500, the differences between the models are minimal. These visual findings align with the numerical summaries and confirm that the Bayesian model provides more robust interval estimation under limited data conditions, especially for smaller or more variable subgroups.

## Results: Shannon Diversity

The second analysis focuses on the Shannon diversity as the response variable, modeled using a Gaussian likelihood. The analysis follows the same simulation and modeling framework as with ARG load: five independent subsamples were drawn for each of the sample sizes (30, 80, 150, 250, 500), and both Bayesian and frequentist models were fitted to each. The aim is again to compare the estimated effects of age categories and evaluate the width and variability of the resulting 95% uncertainty intervals across varying sample sizes.

```{r}
# Plot the distribution of 95% interval widths by model and sample size (Shannon model) 
# Separate panels are shown for each age category
results_shannon$Predictor <- factor(results_shannon$Predictor, levels = desired_order)
ggplot(results_shannon, aes(x = factor(sample_size), y = width, fill = model)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.85, linewidth = 0.6) +
  facet_wrap(~Predictor, scales = "free_y", ncol = 3) +
  labs(title = "Shannon Index: 95% Interval Width by Sample Size",
       x = "Sample Size", y = "95% Interval Width",
       fill = "Model") +
  scale_fill_manual(values = c("Bayesian" = "green",
                               "Frequentist" = "coral")) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face = "bold")
  )


# Create a summary table with mean and standard deviation of interval widths
# Grouped by model type and sample size (across all age categories)
summary_table_shannon <- results_shannon %>%
  group_by(model, sample_size) %>%
  summarise(
    mean_width = mean(width),
    sd_width = sd(width),
    .groups = "drop"
  ) %>%
  arrange(sample_size, model)

summary_table_shannon

```

While the overall pattern for the Shannon diversity index is similar to that observed with ARG load, the convergence between the Bayesian and frequentist models appears to occur slightly faster. At smaller sample sizes (e.g., n = 30 and 80), the Bayesian intervals are again somewhat narrower and more stable, but the difference diminishes more quickly as the sample size increases. By n = 150, the average interval widths are already very close. This may reflect the more symmetric distribution of the Shannon index, which is modeled using a Gaussian likelihood rather than a lognormal distribution.

# Enrichment model

To investigate enrichment patterns in binary outcomes across discrete population groups, we used Bayesian regression models with a Bernoulli likelihood. This approach allows us to estimate the population frequency of a binary event such as the presence of a high ARG load across categories like age group or gender, instead of modeling continuous variation.

By converting continuous outcomes into a binary form for example, defining the top 10% quantile as “high”, we can focus on the enrichment or overrepresentation of extreme values in specific subpopulations. This simple yet powerful framework helps identify disproportionate effects that may be masked when using standard Gaussian models.

For each subgroup, we fit a model with the group variable as a predictor, enabling us to estimate the posterior prevalence and credible intervals for the binary outcome across categories. Enrichment model can be applied across different stratifications, such as by income level, gender, or age, to uncover patterns of enrichment in the population.

Model fitting is computationally intensive and has been performed in a separate R scripts (`EnrichmentModels.R`). Please ensure that this script is available in your working directory.

The resulting model objects are saved as .rds files for each model. These files are then loaded into this document to allow downstream summarization and visualization without re-running the computationally intensive model fitting.

## Grouping by age cateogry and income group

```{r, include=FALSE}
# Load saved Bayesian model fits
fit_enrichment_age <- readRDS("fit_enrichment_age.rds")
# Load enrichment data
enrichment_df <- readRDS("enrichment_df.rds")
```

We use a stratified Bayesian logistic regression model to estimate the prevalence of high ARG load across age categories. Separate models are fitted for each income group (LMIC, HIC). Age category is included as a categorical predictor, which estimates the difference between each age group.

We use a Bernoulli likelihood to model the binary outcome:

$$ 
y_{ij} \sim \text{Bernoulli}(p_{ij}) 
$$

$$
\text{logit}(p_{ij}) = \beta_{0j} + \sum_l \beta_{lj} \cdot \text{AgeCategory}_{ijl}
$$

Where:

-   $y_{ij}$ is the binary outcome (high ARG load) for individual $i$ in income group $j$\
-   $p_{ij}$ is the probability of high ARG load for individual $i$ in income group $j$\
-   $\beta_{0j}$ is the intercept for income group $j$\
-   $\beta_{lj}$ are the fixed effects for each age category $l$ in income group $j$\
-   $\text{AgeCategory}_{ijl}$ is a dummy variable (1 if individual $i$ is in category $l$, 0 otherwise)

We extract posterior summaries (posterior means and 95% credible intervals) of predicted outcome prevalence from each model. Specifically, we generate posterior predictions for each group, summarize these predictions across posterior draws, and compile the results into a combined summary table. This table includes the estimated prevalence, uncertainty intervals, and relevant grouping variables. To support clear visualization, we also clean and standardize column names and ensure that category ordering is consistent across plots.

```{r}
# Create posterior summary table using lapply
summary_df <- lapply(names(fit_enrichment_age), function(ic) {
  
  # Create prediction data
  newdata <- enrichment_df %>%
    filter(income_group == ic) %>%
    distinct(age_category) %>%
    arrange(age_category)
  
  # Posterior expected prevalence
  posterior_preds <- posterior_epred(fit_enrichment_age[[ic]], newdata = newdata)
  
  # Summarize posterior predictions
  summary_ic <- as.data.frame(t(apply(posterior_preds, 2, function(x) {
    c(Estimate = mean(x), Q2_5 = quantile(x, 0.025), Q97_5 = quantile(x, 0.975))
  })))
  
  # Add metadata
  summary_ic$age_category <- newdata$age_category
  summary_ic$income_group <- ic
  
  return(summary_ic)
  
}) %>% bind_rows()

# Clean column names
colnames(summary_df)[1:3] <- c("Estimate", "Q2_5", "Q97_5")

```

Next, we can visualize the results.

```{r}
# Plot
ggplot(summary_df, aes(x = age_category, y = Estimate * 100, color = income_group, group = income_group)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Q2_5 * 100, ymax = Q97_5 * 100), width = 0.2) +
  geom_line(linewidth = 0.5) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "gray40") +
  facet_wrap(~ income_group, ncol = 1) +
  scale_color_manual(values = c("LMIC" = "orange", "HIC" = "blue")) +
  labs(
    x = "Age Category",
    y = "Estimated High ARG Prevalence (%)",
    title = "High ARG Load Enrichment by Age Group and Income Level",
    color = "Income Group"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )


```

In both HIC and LMIC populations, infants show similarly high prevalence of high ARG load. In HICs, prevalence drops after infancy and then gradually rises again in older age groups. In contrast, LMICs show less of a drop after infancy and a much steeper increase in prevalence in later adulthood. This suggests that while both settings share a pattern of early-life elevation, the accumulation of ARG burden with age is much more pronounced in LMIC populations.

## Grouping by gender and income group

```{r, include=FALSE}
# Load saved Bayesian model fits
fit_enrichment_gender <- readRDS("fit_enrichment_gender.rds")
```

We use a stratified Bayesian logistic regression model to estimate the prevalence of high ARG load across gender groups. Separate models are fitted for each income group (LMIC, HIC). Gender is included as a categorical predictor, which estimates the difference between genders.

We use a Bernoulli likelihood to model the binary outcome:

$$ 
y_{ij} \sim \text{Bernoulli}(p_{ij}) 
$$

$$
\text{logit}(p_{ij}) = \beta_{0j} + \sum_m \beta_{mj} \cdot \text{Gender}_{ijm}
$$

Where:

- $y_{ij}$ is the binary outcome (high ARG load) for individual $i$ in income group $j$  
- $p_{ij}$ is the probability of high ARG load for individual $i$ in income group $j$  
- $\beta_{0j}$ is the intercept for income group $j$  
- $\beta_{mj}$ are the fixed effects for each gender $m$ in income group $j$  
- $\text{Gender}_{ijm}$ is a dummy variable (1 if individual $i$ is in gender category $m$, 0 otherwise)

We repeat the same procedure as above, extracting posterior means and 95% credible intervals of predicted prevalence, summarizing by group, and preparing the results for visualization.

```{r}
# Create posterior summary table using lapply
summary_df <- lapply(names(fit_enrichment_gender), function(ic) {
  
  # Create prediction data
  newdata <- enrichment_df %>%
    filter(income_group == ic) %>%
    distinct(gender) %>%
    arrange(gender)
  
  # Posterior expected prevalence
  posterior_preds <- posterior_epred(fit_enrichment_gender[[ic]], newdata = newdata)
  
  # Summarize posterior predictions
  summary_ic <- as.data.frame(t(apply(posterior_preds, 2, function(x) {
    c(Estimate = mean(x), Q2_5 = quantile(x, 0.025), Q97_5 = quantile(x, 0.975))
  })))
  
  # Add metadata
  summary_ic$gender <- newdata$gender
  summary_ic$income_group <- ic
  
  return(summary_ic)
  
}) %>% bind_rows()

# Clean column names
colnames(summary_df)[1:3] <- c("Estimate", "Q2_5", "Q97_5")
```

Next, we visualize the results.

```{r}
# Plot
ggplot(summary_df, aes(x = gender, y = Estimate * 100, color = income_group, group = income_group)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Q2_5 * 100, ymax = Q97_5 * 100), width = 0.2) +
  geom_line(linewidth = 0.5) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "gray40") +
  facet_wrap(~ income_group, ncol = 1) +
  scale_color_manual(values = c("LMIC" = "orange", "HIC" = "blue")) +
  labs(
    x = "Gender",
    y = "Estimated High ARG Prevalence (%)",
    title = "High ARG Load Enrichment by Gender and Income Level",
    color = "Income Group"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )

```

In both HIC and LMIC populations, women show slightly higher estimated prevalence of high ARG load compared to men, but the differences are not statistically significant, as the credible intervals overlap. The overall prevalence is somewhat lower in HICs and higher in LMICs, with greater uncertainty in the LMIC estimates.

## Grouping by conitnent

```{r, include=FALSE}
# Load saved Bayesian model fit
fit_enrichment_continent <- readRDS("fit_enrichment_continent.rds")
```

We use a Bayesian logistic regression model to estimate the prevalence of high ARG load across continents. A single model is fitted using all data, with continent included as a categorical predictor that estimates the difference between continents.

We use a Bernoulli likelihood to model the binary outcome:

$$ 
y_i \sim \text{Bernoulli}(p_i) 
$$

$$
\text{logit}(p_i) = \beta_0 + \sum_c \beta_c \cdot \text{Continent}_{ic}
$$

Where:

- $y_i$ is the binary outcome (high ARG load) for individual $i$  
- $p_i$ is the probability of high ARG load for individual $i$  
- $\beta_0$ is the overall intercept  
- $\beta_c$ are the fixed effects for each continent $c$  
- $\text{Continent}_{ic}$ is a dummy variable (1 if individual $i$ is in continent $c$, 0 otherwise)

We repeat the same procedure as above, extracting posterior means and 95% credible intervals of predicted prevalence, summarizing by group, and preparing the results for visualization.

```{r}
# Create posterior summary table
summary_df <- {
  # Get allowed levels from model data
  allowed_levels <- levels(fit_enrichment_continent[["all"]]$data$continent)
  
  # Get prediction data and match factor levels
  newdata <- enrichment_df %>%
    distinct(continent) %>%
    filter(continent %in% allowed_levels) %>%
    arrange(continent)
  
  # Make sure continent is a factor with allowed levels
  newdata$continent <- factor(newdata$continent, levels = allowed_levels)
  
  # Posterior expected prevalence
  posterior_preds <- posterior_epred(fit_enrichment_continent[["all"]], newdata = newdata)
  
  # Summarize posterior predictions
  summary_ic <- as.data.frame(t(apply(posterior_preds, 2, function(x) {
    c(Estimate = mean(x), Q2_5 = quantile(x, 0.025), Q97_5 = quantile(x, 0.975))
  })))
  
  # Add metadata
  summary_ic$continent <- newdata$continent
  
  # Clean column names
  colnames(summary_ic)[1:3] <- c("Estimate", "Q2_5", "Q97_5")
  
  summary_ic
}
```

Next, we visualize the results.

```{r}
# Plot
ggplot(summary_df, aes(x = continent, y = Estimate * 100)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Q2_5 * 100, ymax = Q97_5 * 100), width = 0.2) +
  geom_line(aes(group = 1), linewidth = 0.5) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "gray40") +
  labs(
    x = "Continent",
    y = "Estimated High ARG Prevalence (%)",
    title = "High ARG Load Enrichment by Continent"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )

```

The estimated prevalence of high ARG load varies across continents. Asia and Europe show the highest estimates, with Asia having particularly wide uncertainty due to the small number of observations. North America shows intermediate levels, while Oceania has the lowest estimate.



## Hierarchical enrichment model

### Age catecory and income group

```{r, include=FALSE}
# Load saved hierarchical model
fit_enrichment_age_hier <- readRDS("fit_enrichment_age_hier.rds")
```


We use a hierarchical Bayesian logistic regression model to estimate the prevalence of high ARG load across age categories, allowing the effects of age to vary across income groups. A single model is fitted using all data, with age category as a categorical predictor and random intercepts and slopes by income group.

We use a Bernoulli likelihood to model the binary outcome:

$$ 
y_{ij} \sim \text{Bernoulli}(p_{ij}) 
$$

$$
\text{logit}(p_{ij}) = \beta_0 + \sum_l \beta_l \cdot \text{AgeCategory}_{ijl} + b_{0j} + \sum_l b_{lj} \cdot \text{AgeCategory}_{ijl}
$$

Where:

- $y_{ij}$ is the binary outcome (high ARG load) for individual $i$ in income group $j$  
- $p_{ij}$ is the probability of high ARG load for individual $i$ in income group $j$  
- $\beta_0$ is the overall intercept  
- $\beta_l$ are the fixed effects for each age category $l$  
- $b_{0j}$ is the random intercept for income group $j$  
- $b_{lj}$ are the random slopes for age category $l$ within income group $j$  
- $\text{AgeCategory}_{ijl}$ is a dummy variable (1 if individual $i$ is in age category $l$, 0 otherwise)

We repeat the same procedure as above, extracting posterior means and 95% credible intervals of predicted prevalence, summarizing by group, and preparing the results for visualization.

```{r}
# Create prediction data across all income groups
newdata <- enrichment_df %>%
  distinct(age_category, income_group) %>%
  arrange(income_group, age_category)

# Posterior expected prevalence
posterior_preds <- posterior_epred(fit_enrichment_age_hier[["all"]], newdata = newdata)

# Summarize posterior predictions
summary_df <- as.data.frame(t(apply(posterior_preds, 2, function(x) {
  c(Estimate = mean(x), Q2_5 = quantile(x, 0.025), Q97_5 = quantile(x, 0.975))
})))

# Add metadata
summary_df$age_category <- newdata$age_category
summary_df$income_group <- newdata$income_group

# Clean column names
colnames(summary_df)[1:3] <- c("Estimate", "Q2_5", "Q97_5")

```


Next we visualize the results.


```{r}
# Plot
ggplot(summary_df, aes(x = age_category, y = Estimate * 100, color = income_group, group = income_group)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Q2_5 * 100, ymax = Q97_5 * 100), width = 0.2) +
  geom_line(linewidth = 0.5) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "gray40") +
  facet_wrap(~ income_group, ncol = 1) +
  scale_color_manual(values = c("LMIC" = "orange", "HIC" = "blue")) +
  labs(
    x = "Age Category",
    y = "Estimated High ARG Prevalence (%)",
    title = "High ARG Load Enrichment by Age Group and Income Level (Hierarchical Model)",
    color = "Income Group"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )

```

The hierarchical model produces results that are broadly consistent with the simpler stratified age-category model. The estimated patterns of high ARG prevalence across age groups and income levels remain similar, indicating that these enrichment patterns are robust to modeling approach.


### Gender and income group


```{r, include=FALSE}
# Load saved hierarchical model
fit_enrichment_gender_hier <- readRDS("fit_enrichment_gender_hier.rds")
```

We use a hierarchical Bayesian logistic regression model to estimate the prevalence of high ARG load across gender groups, allowing the effects of gender to vary across income groups. A single model is fitted using all data, with gender as a categorical predictor and random intercepts and slopes by income group.

We use a Bernoulli likelihood to model the binary outcome:

$$ 
y_{ij} \sim \text{Bernoulli}(p_{ij}) 
$$

$$
\text{logit}(p_{ij}) = \beta_0 + \sum_m \beta_m \cdot \text{Gender}_{ijm} + b_{0j} + \sum_m b_{mj} \cdot \text{Gender}_{ijm}
$$

Where:

- $y_{ij}$ is the binary outcome (high ARG load) for individual $i$ in income group $j$  
- $p_{ij}$ is the probability of high ARG load for individual $i$ in income group $j$  
- $\beta_0$ is the overall intercept  
- $\beta_m$ are the fixed effects for each gender $m$  
- $b_{0j}$ is the random intercept for income group $j$  
- $b_{mj}$ are the random slopes for gender $m$ within income group $j$  
- $\text{Gender}_{ijm}$ is a dummy variable (1 if individual $i$ is in gender category $m$, 0 otherwise)

We repeat the same procedure as above, extracting posterior means and 95% credible intervals of predicted prevalence, summarizing by group, and preparing the results for visualization.

```{r}
# Create prediction data across all income groups
newdata <- enrichment_df %>%
  distinct(gender, income_group) %>%
  arrange(income_group, gender)

# Posterior expected prevalence
posterior_preds <- posterior_epred(fit_enrichment_gender_hier[["all"]], newdata = newdata)

# Summarize posterior predictions
summary_df <- as.data.frame(t(apply(posterior_preds, 2, function(x) {
  c(Estimate = mean(x), Q2_5 = quantile(x, 0.025), Q97_5 = quantile(x, 0.975))
})))

# Add metadata
summary_df$gender <- newdata$gender
summary_df$income_group <- newdata$income_group

# Clean column names
colnames(summary_df)[1:3] <- c("Estimate", "Q2_5", "Q97_5")
```


Next we visualize the results.

```{r}
# Plot
ggplot(summary_df, aes(x = gender, y = Estimate * 100, color = income_group, group = income_group)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Q2_5 * 100, ymax = Q97_5 * 100), width = 0.2) +
  geom_line(linewidth = 0.5) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "gray40") +
  facet_wrap(~ income_group, ncol = 1) +
  scale_color_manual(values = c("LMIC" = "orange", "HIC" = "blue")) +
  labs(
    x = "Gender",
    y = "Estimated High ARG Prevalence (%)",
    title = "High ARG Load Enrichment by Gender and Income Level (Hierarchical Model)",
    color = "Income Group"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
```


The hierarchical model produces patterns that are very similar to the original stratified gender model. The estimated high ARG load prevalence across gender and income levels shows consistent trends, suggesting that the main findings are robust regardless of whether a stratified or hierarchical modeling approach is used.




### Gender, age category and income group

```{r, include=FALSE}
# Load saved hierarchical model
fit_enrichment_age_gender_hier <- readRDS("fit_enrichment_age_gender_hier.rds")
```

We use a hierarchical Bayesian logistic regression model to estimate the prevalence of high ARG load across age categories, allowing the effects of age to vary across combinations of income group and gender. A single model is fitted using all data, with age category as a categorical predictor and random intercepts and slopes by income group and gender.

We use a Bernoulli likelihood to model the binary outcome:

$$ 
y_{ijm} \sim \text{Bernoulli}(p_{ijm}) 
$$

$$
\text{logit}(p_{ijm}) = \beta_0 + \sum_l \beta_l \cdot \text{AgeCategory}_{ijm l} + b_{0 j m} + \sum_l b_{l j m} \cdot \text{AgeCategory}_{ijm l}
$$

Where:

- $y_{ijm}$ = binary outcome (high ARG load) for individual $i$ in income group $j$ and gender $m$  
- $p_{ijm}$ = probability of high ARG load for individual $i$ in that group  
- $\beta_0$ = overall intercept  
- $\beta_l$ = fixed effects for each age category $l$  
- $b_{0 j m}$ = random intercept for income group $j$ and gender $m$  
- $b_{l j m}$ = random slope for age category $l$ within income group $j$ and gender $m$  
- $\text{AgeCategory}_{ijm l}$ = dummy variable (1 if individual $i$ is in category $l$, 0 otherwise)

We extract posterior summaries as before and prepare the results for visualization.

```{r}
# Create prediction data across all income and gender groups
newdata <- enrichment_df %>%
  distinct(age_category, income_group, gender) %>%
  arrange(income_group, gender, age_category)

# Posterior expected prevalence
posterior_preds <- posterior_epred(fit_enrichment_age_gender_hier[["all"]], newdata = newdata)

# Summarize posterior predictions
summary_df <- as.data.frame(t(apply(posterior_preds, 2, function(x) {
  c(Estimate = mean(x), Q2_5 = quantile(x, 0.025), Q97_5 = quantile(x, 0.975))
})))

# Add metadata
summary_df$age_category <- newdata$age_category
summary_df$income_group <- newdata$income_group
summary_df$gender <- newdata$gender

# Clean column names
colnames(summary_df)[1:3] <- c("Estimate", "Q2_5", "Q97_5")
```

Next we visualize the results.

```{r, fig.width=14, fig.height=10}
# Plot
ggplot(summary_df, aes(x = age_category, y = Estimate * 100, color = gender, group = gender)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Q2_5 * 100, ymax = Q97_5 * 100), width = 0.2) +
  geom_line(linewidth = 0.5) +
  facet_wrap(~ income_group, ncol = 1) +
  scale_color_manual(values = c("Women" = "red", "Men" = "blue")) +
  labs(
    x = "Age Category",
    y = "Estimated High ARG Prevalence (%)",
    title = "High ARG Load Enrichment by Age, Gender, and Income Level (Hierarchical Model)",
    color = "Gender"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

The hierarchical model with random slopes shows that age-related patterns vary across income groups, with a steeper increase in prevalence with age in LMICs, while HIC patterns remain relatively stable. Credible intervals mainly overlap across groups, indicating no clear differences between genders.


### Continent and age category

```{r, include=FALSE}
# Load saved hierarchical model
fit_enrichment_age_continent_hier <- readRDS("fit_enrichment_age_continent_hier.rds")
```

We use a hierarchical Bayesian logistic regression model to estimate the prevalence of high ARG load across age categories, allowing the effects of age to vary across continents. This model includes both random intercepts and random slopes for age category, providing flexibility to capture continent-specific age patterns.

We use a Bernoulli likelihood to model the binary outcome:

$$ 
y_{i c} \sim \text{Bernoulli}(p_{i c}) 
$$

$$
\text{logit}(p_{i c}) = \beta_0 + \sum_l \beta_l \cdot \text{AgeCategory}_{i c l} + b_{0 c} + \sum_l b_{l c} \cdot \text{AgeCategory}_{i c l}
$$

Where:

- $y_{i c}$ = binary outcome (high ARG load) for individual $i$ in continent $c$  
- $p_{i c}$ = probability of high ARG load for individual $i$ in that continent  
- $\beta_0$ = overall intercept  
- $\beta_l$ = fixed effects for each age category $l$  
- $b_{0 c}$ = random intercept for continent $c$  
- $b_{l c}$ = random slope for age category $l$ within continent $c$  
- $\text{AgeCategory}_{i c l}$ = dummy variable (1 if individual $i$ is in age category $l$, 0 otherwise)

We extract posterior summaries as before and prepare the results for visualization.

```{r}
# Create prediction data across all combinations
newdata <- enrichment_df %>%
  distinct(age_category, continent) %>%
  arrange(continent, age_category)

# Filter out unknown continent levels to match model
valid_continents <- levels(fit_enrichment_age_continent_hier[["all"]]$data$continent)
newdata <- newdata %>%
  filter(continent %in% valid_continents)
newdata$continent <- factor(newdata$continent, levels = valid_continents)


# Posterior expected prevalence
posterior_preds <- posterior_epred(fit_enrichment_age_continent_hier[["all"]], newdata = newdata)

# Summarize posterior predictions
summary_df <- as.data.frame(t(apply(posterior_preds, 2, function(x) {
  c(Estimate = mean(x), Q2_5 = quantile(x, 0.025), Q97_5 = quantile(x, 0.975))
})))

# Add metadata
summary_df$age_category <- newdata$age_category
summary_df$continent <- newdata$continent

# Clean column names
colnames(summary_df)[1:3] <- c("Estimate", "Q2_5", "Q97_5")
```

Next we visualize the results.

```{r, fig.width=14, fig.height=10}
# Plot
ggplot(summary_df, aes(x = age_category, y = Estimate * 100, group = continent)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Q2_5 * 100, ymax = Q97_5 * 100), width = 0.1) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ continent, ncol = 1) +
  labs(
    x = "Age Category",
    y = "Estimated High ARG Prevalence (%)",
    title = "High ARG Load Enrichment by Age and Continent (Hierarchical Model with Slopes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

The figure shows continent-specific age patterns in estimated high ARG prevalence. Asia and Europe show higher prevalence levels, especially in younger age groups. 




### Continent, gender, age category and income group (no slopes)

```{r, include=FALSE}
# Load saved hierarchical model
fit_enrichment_income_age_gender_continent_hier <- readRDS("fit_enrichment_income_age_gender_continent_hier.rds")
```

We use a hierarchical Bayesian logistic regression model to estimate the prevalence of high ARG load across age categories, with random intercepts by continent, income group, and gender. This simplified model retains age category as a categorical predictor but includes only random intercepts (no random slopes), making the model faster to fit.

We use a Bernoulli likelihood to model the binary outcome:

$$ 
y_{ijm c} \sim \text{Bernoulli}(p_{ijm c}) 
$$

$$
\text{logit}(p_{ijm c}) = \beta_0 + \sum_l \beta_l \cdot \text{AgeCategory}_{ijm c l} + b_{0 j m c}
$$

Where:

- $y_{ijm c}$ = binary outcome (high ARG load) for individual $i$ in income group $j$, gender $m$, continent $c$  
- $p_{ijm c}$ = probability of high ARG load for individual $i$ in that group  
- $\beta_0$ = overall intercept  
- $\beta_l$ = fixed effects for each age category $l$  
- $b_{0 j m c}$ = random intercept for income group $j$, gender $m$, continent $c$  
- $\text{AgeCategory}_{ijm c l}$ = dummy variable (1 if individual $i$ is in age category $l$, 0 otherwise)

We extract posterior summaries as before and prepare the results for visualization.

```{r}
# Create prediction data across all combinations
newdata <- enrichment_df %>%
  distinct(age_category, income_group, gender, continent) %>%
  arrange(continent, income_group, gender, age_category)

# Filter out unknown continent levels to match model
valid_continents <- levels(fit_enrichment_income_age_gender_continent_hier[["all"]]$data$continent)
newdata <- newdata %>%
  filter(continent %in% valid_continents)
newdata$continent <- factor(newdata$continent, levels = valid_continents)

# Posterior expected prevalence
posterior_preds <- posterior_epred(
  fit_enrichment_income_age_gender_continent_hier[["all"]],
  newdata = newdata
)

# Summarize posterior predictions
summary_df <- as.data.frame(t(apply(posterior_preds, 2, function(x) {
  c(Estimate = mean(x), Q2_5 = quantile(x, 0.025), Q97_5 = quantile(x, 0.975))
})))

# Add metadata
summary_df$age_category <- newdata$age_category
summary_df$income_group <- newdata$income_group
summary_df$gender <- newdata$gender
summary_df$continent <- newdata$continent

# Clean column names
colnames(summary_df)[1:3] <- c("Estimate", "Q2_5", "Q97_5")
```

Next we visualize the results.

```{r, fig.width=14, fig.height=10}
# Plot
ggplot(summary_df, aes(x = age_category, y = Estimate * 100, color = gender, group = gender)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Q2_5 * 100, ymax = Q97_5 * 100), width = 0.2) +
  geom_line(linewidth = 0.5) +
  facet_grid(continent ~ income_group) +
  scale_color_manual(values = c("Women" = "red", "Men" = "blue")) +
  labs(
    x = "Age Category",
    y = "Estimated High ARG Prevalence (%)",
    title = "High ARG Load Enrichment by Age, Gender, Income Level, and Continent (Hierarchical Model, Intercepts Only)",
    color = "Gender"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

The intercept-only model shows fairly parallel age trends across gender, income group, and continent, with women slightly higher on average but overlapping credible intervals. Next, we will fit a model with random slopes to explore whether age effects differ more strongly between groups.




### Continent, income group, gender, and age (with slopes)

```{r, include=FALSE}
# Load saved hierarchical model
fit_enrichment_all_hier <- readRDS("fit_enrichment_all_hier.rds")
```

We use a hierarchical Bayesian logistic regression model to estimate the prevalence of high ARG load across age categories, allowing the effects of age to vary across combinations of continent, income group, and gender. This extended model includes both random intercepts and random slopes for age category, providing maximum flexibility to capture group-specific age patterns.

We use a Bernoulli likelihood to model the binary outcome:

$$ 
y_{ijmc} \sim \text{Bernoulli}(p_{ijmc}) 
$$

$$
\text{logit}(p_{ijmc}) = \beta_0 + \sum_l \beta_l \cdot \text{AgeCategory}_{ijmcl} + b_{0jmc} + \sum_l b_{ljmc} \cdot \text{AgeCategory}_{ijmcl}
$$

Where:

- $y_{ijmc}$ = binary outcome (high ARG load) for individual $i$ in income group $j$, gender $m$, continent $c$  
- $p_{ijmc}$ = probability of high ARG load for individual $i$ in that group  
- $\beta_0$ = overall intercept  
- $\beta_l$ = fixed effects for each age category $l$  
- $b_{0jmc}$ = random intercept for income group $j$, gender $m$, continent $c$  
- $b_{ljmc}$ = random slope for age category $l$ within income group $j$, gender $m$, continent $c$  
- $\text{AgeCategory}_{ijmcl}$ = dummy variable (1 if individual $i$ is in age category $l$, 0 otherwise)

We extract posterior summaries as before and prepare the results for visualization.

```{r}
# Create prediction data across all combinations
newdata <- enrichment_df %>%
  distinct(age_category, income_group, gender, continent) %>%
  arrange(continent, income_group, gender, age_category)

# Filter out unknown continent levels to match model
valid_continents <- levels(fit_enrichment_all_hier[["all"]]$data$continent)
newdata <- newdata %>%
  filter(continent %in% valid_continents)
newdata$continent <- factor(newdata$continent, levels = valid_continents)

# Posterior expected prevalence
posterior_preds <- posterior_epred(fit_enrichment_all_hier[["all"]], newdata = newdata)

# Summarize posterior predictions
summary_df <- as.data.frame(t(apply(posterior_preds, 2, function(x) {
  c(Estimate = mean(x), Q2_5 = quantile(x, 0.025), Q97_5 = quantile(x, 0.975))
})))

# Add metadata
summary_df$age_category <- newdata$age_category
summary_df$income_group <- newdata$income_group
summary_df$gender <- newdata$gender
summary_df$continent <- newdata$continent

# Clean column names
colnames(summary_df)[1:3] <- c("Estimate", "Q2_5", "Q97_5")
```

Next we visualize the results.

```{r, fig.width=14, fig.height=10}
# Plot
ggplot(summary_df, aes(x = age_category, y = Estimate * 100, color = gender, group = gender)) +
  geom_point(size = 1.5) +
  geom_errorbar(aes(ymin = Q2_5 * 100, ymax = Q97_5 * 100), width = 0.1) +
  geom_line(linewidth = 0.6) +
  facet_grid(continent ~ income_group) +
  scale_color_manual(values = c("Women" = "red", "Men" = "blue")) +
  labs(
    x = "Age Category",
    y = "Estimated High ARG Prevalence (%)",
    title = "High ARG Load Enrichment by Age, Gender, Income Level, and Continent (Hierarchical Model with Slopes)",
    color = "Gender"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

The random slope model reveals variation in age trends across gender, income group, and continent. While credible intervals remain wide, especially in LMICs and some continents, overall patterns do not show pronounced gender differences.



# Comparing rstan and cmdstanr

To assess the computational performance of different Stan backends, we compared model fitting times using both rstan and cmdstanr. Two statistical models were tested: Model 4, which includes a random intercept for bioproject and accounts for sequencing depth, and a hierarchical enrichment model, which estimates the prevalence of high ARG load across gender groups with varying effects across income groups. Both models were previously fitted using rstan earlier in the analysis. Each model was fitted five times with both backends to obtain reliable timing estimates.

This benchmarking was conducted in a separate R script (`StanCompare.R`). Please ensure that this script is available in your working directory. The results have been saved as `.rds` file and are loaded here to enable downstream analysis.

```{r, include=FALSE}
# Load results
stan_timings <- readRDS("stan_timing_results.rds")
```


To summarize backend performance, we calculated the average model fitting time for each Stan backend across three repetitions. After that we visualize the results.

```{r}
# Compute mean timing per model and backend
avg_timings <- stan_timings %>%
  group_by(model, backend) %>%
  summarise(mean_time_sec = mean(time_sec), .groups = "drop")

# Print summary table
print(avg_timings)

# Plot: mean runtime by model and backend
ggplot(avg_timings, aes(x = model, y = mean_time_sec, fill = backend)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  labs(
    title = "Mean model fitting time by Stan backend",
    x = "Model",
    y = "Mean time (seconds)",
    fill = "Backend"
  ) +
  theme_minimal(base_size = 12)
```


Cmdstanr consistently outperformed rstan in terms of model fitting time. The performance gap was especially notable for the hierarchical enrichment model, where cmdstanr completed inference in nearly half the time required by rstan.